---
title: "STAT 3400 - Homework 6"
author: "Taketo Horigome"
date: "Due March 1, 2026"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(tinytex)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%")
```


# Problem 1

There are two common representations of a simple logistic regression model. The first is in the probability form shown below.

$$
P(y=1|x)=\frac{1}{1+e^{-(\beta_0+\beta_1x)}}
$$

The second representation is the log-odds form shown below.

$$
\log\left(\frac{P(y=1|x)}{P(y=0|x)}\right)=\beta_0+\beta_1x
$$

Using some algebra, show that these two representations are mathematically equivalent.

#### Solution:

Let's start from $\log\left(\frac{P(y=1|x)}{P(y=0|x)}\right)=\beta_0+\beta_1x$.

Exponentiate both sides:  $\left(\frac{P(y=1|x)}{P(y=0|x)}\right) = e^{\beta_0+\beta_1x}$.

Since this is binary classification, let $p = {P(y=1|x)}$, then ${P(y=0|x)} = 1-p$, so we can write the equation as $\frac{p}{1-p} = e^{\beta_0+\beta_1x}$.

Solve for p by multiplying both sides by $1-p$: $p = e^{\beta_0+\beta_1x}(1-p) = e^{\beta_0+\beta_1x} - e^{\beta_0+\beta_1x}p$.

We can factor out p: $p(1 + e^{\beta_0+\beta_1x}) = e^{\beta_0+\beta_1x}$, then we can solve for p: $p = \frac{e^{\beta_0+\beta_1x}}{1 + e^{\beta_0+\beta_1x}}$.

Finally if we multiply numerator and denominator by $e^{-(\beta_0+\beta_1x)}$: $p = \frac{1}{1 + e^{-(\beta_0+\beta_1x)}}$.

Since $p = P(y=1|x)$, we have $P(y=1|x)=\frac{1}{1+e^{-(\beta_0+\beta_1x)}}$.




# Problem 2

This problem regards the conversion between various forms of representing the likelihood of an event. Specifically, the problem involves a binary response variable that is equal to 1 when an individual defaults on their credit card debt and 0 when they do not default.

## Part A

On average, what fraction of people with an odds of 0.37 of defaulting on their credit card debt will in fact default?  Show your work.

#### Solution:

Let p be the probability that a person defaults.
Then, $0.37 = \frac{p}{1-p}$.
Which is equivalent to $0.37*(1-p) = p$, $0.37 = 1.37p$.
Thus, $p = \frac{0.37}{1.37} = 0.27....$
Therefore, about 27% of people with odds 0.37 of defaulting will actually default.

## Part B

Suppose that an individual has a 16% chance of defaulting on their credit card debt. What are the odds that the individual will default? Show your work.

#### Solution:

In this case, since the $p = 0.16$, the odds are $\frac{0.16}{1-0.16} = \frac{0.16}{0.84} = 0.19...$
Therefore, the odds that individual will default is about 0.19.



# Problem 3

This problem uses the `Auto` data set from the `ISLR2` package.

```{r}
library(ISLR2)
data(Auto)
```

## Part A

Add a new binary variable `mpg_new` that is equal to 1 when `mpg` has a value greater than its median and 0 otherwise. Then, after removing `year`,`origin`, and `name` from the data set, create five separate box plots that show the association between `mpg_new` and the remaining predictors. Which of the predictors appears to be most associated with `mpg_new` based on what you observe in the plots?

```{r}
mpg_median <- median(Auto$mpg, na.rm = TRUE)
Auto$mpg_new <- ifelse(Auto$mpg > mpg_median, 1, 0)
Auto_2 <- select(Auto, -year, -origin, - name)
ggplot(Auto_2, aes(x = factor(mpg_new), y = cylinders)) +
  geom_boxplot() +
  labs(title = "Cylinders vs mpg_new",
       x = "mpg_new",
       y = "Cylinders")

ggplot(Auto_2, aes(x = factor(mpg_new), y = displacement)) +
  geom_boxplot() +
  labs(title = "Displacement vs mpg_new",
       x = "mpg_new",
       y = "Displacement")

ggplot(Auto_2, aes(x = factor(mpg_new), y = horsepower)) +
  geom_boxplot() +
  labs(title = "Horsepower vs mpg_new",
       x = "mpg_new",
       y = "Horsepower")

ggplot(Auto_2, aes(x = factor(mpg_new), y = weight)) +
  geom_boxplot() +
  labs(title = "Weight vs mpg_new",
       x = "mpg_new",
       y = "Weight")
ggplot(Auto_2, aes(x = factor(mpg_new), y = acceleration)) +
  geom_boxplot() +
  labs(title = "Acceleration vs mpg_new",
       x = "mpg_new",
       y = "Acceleration")

```
#### Solution:

From the boxplots, `weight` appears to be most strongly associated with `mpg_new`, as there is a clear separation between cars with `mpg_new = 0` and `mpg_new = 1` and minimal overlap in their distributions. `Horsepower`, `displacement`, and `cylinders` also show strong differences between the two groups. `Acceleration` appears to have the weakest association, as the distributions for `mpg_new = 0` and `mpg_new= 1` overlap considerably. 


## Part B

Using `weight` as the predictor, estimate a simple logistic regression model for the binary response `mpg_new`. Write the equation for your estimated model in *both* probability and log-odds form.

```{r}
model <- glm(mpg_new ~ weight, data = Auto_2, family= "binomial")
summary(model)
```

#### Solution:

From the summary output, we have $\beta_0 = 11.763284$ and $\beta_1 = -0.004119$.
The logistic regression model in log-odds form is: $\log\left(\frac{P(mpg_new = 1|weight)}{1 - P(mpg_new = 1|weight)}\right)=\beta_0+\beta_1x = 11.763284 - 0.004119 * weight$.
The probability form of the model is: ${P(mpg_new = 1 | weight)} = \frac{1}{1+e^{-(\beta_0+\beta_1x)}} = \frac{1}{1 + e^{-(11.763284 - 0.004119 * weight)}}$.

Since the coefficient of weight is negative, as weight increases, the log-odds of having above-median mpg decreases. Therefore, heavier cars are less likely to have high fuel efficiency. 
For every 1-unit increase in weight, the log-odds of having above-median mpg decreases by 0.004119. In terms of odds: $e^{-0.004119} \approx 0.996$.
This means each additional pound multiplies the odds of having high mpg by about 0.996. 


## Part C

Show the six steps of a hypothesis test for the null hypothesis $H_0:\beta_1=0$, where $\beta_1$ is the coefficient for the `weight` predictor in your model from Part B. What does the conclusion of this test mean in the context of a car's weight and fuel efficiency?


#### Solution:

Here are the six steps of a hypothesis: 
-   Define the null and alternate hypotheses
-   Choose a level of significance
-   Compute a test statistic
-   Construct a null distribution
-   Calculate a p-value
-   Decide whether to reject null hypothesis

Step1: Define the null and alternate hypothesis
$H_0: \beta_1 = 0$
$H_a: \beta_1 \neq 0$

Step2: Choose a level of significance: 
Let $\alpha = 0.05$.

Step3: Compute a test statistic:
From the output $\beta_1 = -0.004119$ and $SE(\beta_1) = 0.000413$.
Thus, z-statistic is: $z = \frac{\beta_1 - 0}{SE(\beta_1)} = \frac{-0.004119}{0.000413} \approx -9.974$.

Step 4: Construct a null distribution:
Under $H_0: \beta_1 = 0$, the test statistic approximately follows a standard normal distribution: $z ~ N(0,1)$.

Step5: Calculate a p-value:
Since the p-value is less than $2*10^{-16}$.

Step6: Decision and Conclusion:
Since p-value is less than 0.05, we reject null hypothesis.

Overall, there is very strong evidence that a car's weight is associated with fuel efficiency. Because $\beta_1$ is negative, heavier cars are slightly less likely to have above-median mpg, meaning heavier cars tend to be less fuel efficient. 



# Problem 4

This problem uses the `Boston` data set from the `ISLR2` package.

```{r}
data(Boston)
```

## Part A

Add a new binary variable `crim_new` that is equal to 1 when `crim` has a value greater than its median and 0 otherwise. Then, create four scatter plots with `crim_new` on the $y$-axis and `rm`, `dis`, `ptratio`, and `medv`, respectively, on the $x$-axis. Interpret the observed direction (positive or negative) of association in each plot.  Does this make sense in the context of the data?

```{r}
crim_median <- median(Boston$crim, na.rm = TRUE)
Boston$crim_new <- ifelse(Boston$crim > crim_median, 1, 0)
ggplot(Boston, aes(x = rm, y =crim_new)) +
  geom_jitter(height = 0.05, size = 2, alpha = 0.5) +
  geom_smooth(
    formula = y ~ x, method = "glm",
    method.args = list(family = "binomial"), se = FALSE, color = "blue"
  )+
  labs(title = "rm vs crim_new",
       x = "rm", 
       y = "crim_new")

ggplot(Boston, aes(x = dis, y =crim_new)) +
  geom_jitter(height = 0.05, size = 2, alpha = 0.5) +
  geom_smooth(formula = y~x, method = "glm",
              method.args = list(family = "binomial"), se = FALSE, color= 'blue') +
  labs(title = "dis vs crim_new",
       x = "dis",
       y = "crim_new")

ggplot(Boston, aes(x = ptratio, y = crim_new)) +
  geom_jitter(height = 0.05, size = 2, alpha = 0.5) +
  geom_smooth(formula = y~x, method = "glm",
              method.args = list(family = "binomial"), se = FALSE, color = 'blue') +
  labs(title = "ptratio vs crim_new",
       x = 'ptratio',
       y = 'crim_new')

ggplot(Boston, aes(x = medv, y = crim_new)) +
  geom_jitter(height = 0.05, size = 2, alpha = 0.5) +
  geom_smooth(formula= y~x, method = "glm",
              method.args = list(family = "binomial"), se = FALSE, color = 'blue') +
  labs(title = 'medv vs crim_new',
       x = 'medv',
       y = 'crim_new')
```

#### Solution:

The logistic smooth curves show that `dis` has a strong negative association with `crim_new`, indicating that areas farther from employment centers tend to have lower crime rates. The variable `medv` also shows a strong negative relationship, suggesting that higher home values are associated with lower crime. The variable `ptratio` shows a positive association, meaning that higher pupil-teacher ratios are associated with higher crime. Finally, `rm` shows a moderate negative association, indicating that neighborhoods with more rooms tend to have lower crime. These relationships are consistent with socioeconomic expectations in the Boston data. 

## Part B

Using `medv` as the predictor, estimate a simple logistic regression model for the binary response `crim_new`. Write the equation for your estimated model in *both* probability and log-odds form.

```{r}
boston_model <- glm(crim_new ~ medv, data = Boston, family = "binomial")
summary(boston_model)
```

#### Solution:

From the summary output: $\beta_0 = 1.42773$ and $\beta_1 = -0.06404$.
The logistic regression model in log-odds form is: $\log\left(\frac{P(crim_new = 1|medv)}{1 - P(crim_new = 1|medv)}\right)=\beta_0+\beta_1x = 1.42773 - 0.06404 * medv$.
The probability form of the model is: ${P(crim_new = 1 | medv)} = \frac{1}{1+e^{-(\beta_0+\beta_1x)}} = \frac{1}{1 + e^{-(1.42773 - 0.06404 * medv)}}$.

Since the coefficient of `medv` is negative, as median home value increases, the log-odds of having above-median crime decreases. Therefore, higher home values are associated with lower crime rates.
More specifically, for each one-unit increase in `medv`, the log-odds of high crime decreases by 0.06404. In terms of odds: $e^{-0.06404} \approx 0.938$.
This means each additional unit increase in median home value multiplies the odds of high crime by about 0.938. 

## Part C

Construct a 90% bootstrap confidence interval for the parameter $\beta_1$, where $\beta_1$ is the coefficient for the `medv` predictor in your model from Part B. What do the bounds on this confidence interval suggest in terms of the association between home values and crime rates?

```{r}
set.seed(1)

results2 <- data.frame(beta1 = rep(NA, 2000))

for (i in 1:2000) {
  boot <- Boston[sample(nrow(Boston), replace = TRUE), ]

  remodel2 <- glm(crim_new ~ medv, data = boot, family = "binomial")
  results2$beta1[i] <- coef(remodel2)[2]
}

threshold1 <- as.numeric(quantile(results2$beta1, 0.05))
threshold2 <- as.numeric(quantile(results2$beta1, 0.95))

ggplot(data = results2, aes(x = beta1)) +
  geom_histogram(color = "black", fill = "#CFB87C") +
  geom_vline(
    xintercept = threshold1, color = "red",
    linewidth = 1, linetype = "dashed"
  ) +
  geom_vline(
    xintercept = threshold2, color = "red",
    linewidth = 1, linetype = "dashed"
  ) +
  labs(
    title = "Sampling Distribution",
    subtitle = "90% Confidence Interval",
    x = "Estimated medv Coefficient",
    y = "Count"
  ) +
  theme_bw()
```

```{r}
quantile(results2$beta1, probs = c(0.05, 0.95))
```

#### Solution:

A 90% bootstrap confidence interval for $\beta_1$ is approximately (-0.09596, -0.04241).
Since the entire interval is negative and does not contain 0, this suggests a statistically significant negative association between median home value and crime rate.
Specifically, as median home value increases, the probability of having above-median crime decreases.
