---
title: "STAT 3400 - Homework 7"
author: "YOUR NAME"
date: "Due March 8, 2026"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(tinytex)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%")
```

# Problem 1

Suppose we collect data for a group of students in a statistics class with variables for hours studied ($x_1$), undergraduate GPA ($x_2$), and whether the student earned an A ($y$) in the class. We fit a logistic regression and produce estimated coefficients $\hat\beta_0 = -6.25$, $\hat\beta_1 = 0.1$, and $\hat\beta_2= 0.75$. Reference this model when answering the following questions.

## Part A

Estimate the probability that a student who studies for 40 hours and has an undergraduate GPA of 3.5 gets an A in the class.

### Solution:

We have the multiple logistic model, $P(\text{Grade = A}|x)=\frac{1}{1+e^{-(\beta_0+\beta_1x_1+\beta_2x_2)}}$. We can substitute $x_1 = 40$ and $x_2 = 3.5$ since we want to estimate the probability that a student who studies for 40 hours and has an undergraduate GPA gets an A in the class.$P(\text{Grade = A}|x)=\frac{1}{1 + e^{-(-6.25 + 0.1(40) + 0.75(3.5)}} = \frac{1}{1 + e^{-0.375}} \approx 0.5927$. Thus, the probability that a student who studies for 40 hours and has an undergraduate GPA of 3.5 gets an A in the class based on the given model is about 59.27%.

## Part B

How many hours would the student in Part A need to study to have a 75% chance of getting an A in the class?

### Solution:

We want $\frac{1}{1 + e^{-(-6.25 + 0.1x_1 + 0.753.5)}} = \frac{1}{1 + e^{-(-3.625 + 0.1x_1)}} = 0.75$. Now let use the log transformation: $-3.625 + 0.1x_1 = \log(\frac{0.75}{0.25}) = \log(3) \approx 1.0986$. So, $0.1x_1 = 1.0986 + 3.625 = 4.7236$. $x_1 = \frac{4.7236}{0.1} = 47.236$. Therefore, the student would need to study about 47.24 hours to have a 75% chance of getting an A.

# Problem 2

This question references the `Weekly` data set from the `ISLR2` package. The data includes weekly percentage returns for the S&P 500 stock index between 1990 and 2010. Type `?Weekly` in the Console to obtain further detail about the nine variables in the data set.

```{r}
library(ISLR2)
data(Weekly)
?Weekly
```

## Part A

Produce graphical summaries to explore the association between the direction of market return (up or down) and four of the other variables (one graphic each). Describe the direction and strength of association between each pair of variables in the plots.

```{r}
Weekly2 <- Weekly %>% 
  mutate(Direction = ifelse(Direction == "Up", 1, 0 )) %>%
  na.omit()

ggplot(data = Weekly2, aes(x = Lag1, y = Direction)) +
  geom_jitter(height= 0.05, size = 2, alpha = 0.5)+
  geom_smooth(formula = y ~x,method = 'glm',
               method.args = list(family = 'binomial'), se = FALSE,color = 'blue') +
  labs(title = 'Lag1 vs Direction',
       x = 'Percentage return for previous week',
       y = 'Direction') +
  theme_bw()


ggplot(data = Weekly2, aes(x = Lag2, y = Direction)) +
  geom_jitter(height = 0.05, size = 2, alpha = 0.5) +
  geom_smooth(formula = y ~x, method = 'glm',
              method.args = list(family = 'binomial'), se = FALSE, color = 'blue') +
  labs(title = "Lag2 vs Direction",
       x = 'Percentage return for 2 weeks previous',
       y = 'Direction') +
  theme_bw()

ggplot(data = Weekly2, aes(x = Volume, y = Direction)) +
  geom_jitter(height = 0.05, size = 2, alpha = 0.5) +
  geom_smooth(formula = y ~x, method = 'glm',
              method.args = list(family = 'binomial'), se = FALSE, color = 'blue') +
  labs(title = "Volume vs Direction",
       x = 'Volume of shared traded',
       y = 'Direction') +
  theme_bw()

ggplot(data = Weekly2, aes(x = Today, y = Direction)) +
  geom_jitter(height = 0.05, size = 2, alpha = 0.5) +
  geom_smooth(formula = y ~x, method = 'glm',
              method.args = list(family = 'binomial'), se = FALSE, color = 'blue') +
  labs(title = "Today vs Direction",
       x = 'Percentage return for this week',
       y = 'Direction') +
  theme_bw()

```

### Solution:

I chose `Lag1`, `Lag2`, `Volume`, and `Today` as variables to explore the association between the direction of market return and other predictors. For `Lag1`, the relationship between `Direction` appears to be weak. The observations overlap heavily between the two classes, especially in the range between -5 and 5. The logistic regression line shows a slightly negative slope, which suggests that higher values of `Lag1` are associated with a slightly lower probability of market goes up. However, the relationship is weak and not very informative.

For `Lag2`, the relationship with `Direction` also appears to be weak. Similar to `Lag1`, there is a large overlap between the two classes. The logistic regression line has a weak positive slope, which indicates that larger values of `Lag2` may slightly increase the probability that the market goes up. However, the strength of this association is still weak.

For `Volume`, the relationship with `Direction` is even weaker. The logistic regression line is almost flat, and the points are highly overlapped between the two categories. This suggests that trading volume doesn't have a strong association with whether the market goes up or down.

Finally, the relationship between `Today` and `Direction` is extremely strong. This is expected because `Direction` is directly determined by whether the return for the current week is positive or negative. When the weekly return is positive, the direction is 'Up', and vice versa. Therefore, `Today` perfectly determines `Direction`, and it would not be appropriate to use it as a predictor variable in a model that aims to predict `Direction`.

## Part B

Use the full data set to perform a logistic regression with `Direction` as the response and the five lag variables plus `Volume` as predictors. Use the summary function to print the results. Do any of the predictors appear to be statistically significant? If so, which ones? Does this make intuitive sense in the context of the problem?

```{r}
model <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,data = Weekly, family = 'binomial')
summary(model)
```

### Solution:

For the summary output, `Lag2` has a p-value of 0.0296, which is smaller than 0.05, which means that `Lag2` appears to be statistically significant in predicting the direction of the market return. On the other hand, the other variables all have p-value greater than 0.05, so they don't appear to be statistically significant predictors in this model. However, even though `Lag2` is statistically significant, the result doesn't completely make sense in the real world. It's not very clear why the return from two weeks ago would predict the current week's market direction while the return from one week ago is not significant. Because of this, the significane of `Lag2` may simply be due to random variation rather than a meaningful relationship.

## Part C

Compute the pseudo R-squared for your model in Part B. What does this value mean in the context of the problem. Do you consider this value "good" or "bad" and why?

```{r}
1 - (1486.4 / 1496.2)
```

### Solution:

The pseudo R-squared for the logistic regression model can be calculated as $R^2 = 1 - \frac{\text{Residual Deviance}}{\text{Null Deviance}}$, and we got about 0.065. This value means that the model explains only about 0.65% of the variation in the direction of the market return compared to the null model. In other words, the lag variables and trading volume provide very little improvement over simply predicting the overall average direction of the market. I consider this value as bad and poor, since in the context of this problem, this result suggests that the previous weeks' returns and trading volume don't strongly predict whether the market will go up or down in a given week. This shows the idea of difficulity of the financial market.

## Part D

Estimate a new logistic regression model after removing `Volume` and `Lag5` as predictors. Does this new model appear to perform better? What metric are you using to compare the performance of the two models?

```{r}
model2 <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4,data = Weekly, family = 'binomial')
summary(model2)
```

### Solution:

This model appears to perform slightly better than the previous model because it has a lower AIC value. Since AIC penalizes models for including more variables, it is useful for comparing models with different numbers of predictors. By removing `Volume` and `Lag5`, the model becomes simpler while still explaining the data about as well as the previous model. the lower AIC suggests that this reduced model provides a slightly better balance between model fit and model complexity. Therefore, we use AIC as the metric to compare the performance of the two models, and a model with a lower AIC value is generally preferred because it indicates a better trade-off between goodness of fit and simplicity.

# Problem 3

Using the `Boston` data set from the `ISLR2` library, estimate several multiple logistic regression models to predict whether a given census tract has a crime rate above or below the median. Each of your models should include a different combination of predictor variables. Prior to constructing your models, split the data into 75% training and 25% testing. Which of your models has the lowest classification error rate (CER) against the testing set?

```{r}
data(Boston)
set.seed(45)

Boston2 <- Boston %>%
  mutate(crim = ifelse(crim > median(Boston$crim), 1, 0)) %>%
  na.omit()

rows <- sample(x = nrow(Boston2), size = floor(0.75*nrow(Boston2)), replace = FALSE)
training <- Boston2[rows,]
testing <- Boston2[-rows,]

model_1 <- glm(crim ~ zn + indus + nox, data = training, family = 'binomial')
model_2 <- glm(crim ~ rm + age +dis, data = training, family = 'binomial')
model_3 <- glm(crim ~ tax + ptratio + lstat, data = training,  family = 'binomial')
model_4 <- glm(crim ~ nox + rm + tax + lstat, data = training, family = 'binomial')

pred1 <- predict(model_1, testing, type = 'response')
pred2 <- predict(model_2, testing, type = 'response')
pred3 <- predict(model_3, testing, type = 'response')
pred4 <- predict(model_4, testing, type = 'response')

class1 <- ifelse(pred1 > 0.5, 1, 0)
class2<- ifelse(pred2 > 0.5, 1, 0)
class3 <- ifelse(pred3 > 0.5, 1, 0)
class4 <- ifelse(pred4 > 0.5, 1, 0)

Boston_predict <- 
  tibble(
    CER_1 = mean(class1 != testing$crim),
    CER_2 = mean(class2 != testing$crim),
    CER_3 = mean(class3 != testing$crim),
    CER_4 = mean(class4 != testing$crim)
  )
Boston_predict

```

### Solution:

I made four different logistic models using different combination of predictor variables to classify whether the crime rate in a census tract is above or below the median. The models were trained using 75% of the data, and their performance was evaluated using the remaining 25% testing set. Among the 4 models, Model 4 produced the lowest CER. This model includes the predictors `nox`, `rm`, `tax`, and `lstat`, which appear to provide better predictive performance than the other combination of variables considered. Therefore, the model 4 performs best among the models tested for predicting whether the crime rate is above the median.

# Problem 4

Referring to your best model from Problem 3, construct a confusion matrix from the testing set predictions. How does the false positive rate of your model compare to the false negative rate? How does the false discovery rate of your model compare to the false omission rate? What do these values mean in the context of the problem?

```{r}
conf_mat <- table(Predicted = class4, Actual = testing$crim)
conf_mat
TN <- conf_mat[1,1]
FN <- conf_mat[1,2]
FP <- conf_mat[2,1]
TP <- conf_mat[2,2]
FPR <- FP / (FP + TN)
FPR
FNR <- FN / (FN + TP)
FNR
FDR <- FP / (FP + TP)
FDR
FOR <- FN / (FN + TN)
FOR
```

### Solution:

##### False Positive Rate (FPR) vs False Negative Rate (FNR)

The False Positive Rate measures the proportion of actual negative cases that are incorrectly classified as positive. It is calculated as $\text{FPR} = \frac{\text{False Positive}}{\text{False Positive} + \text{True Negative}} = \frac{12}{12 + 63} = \frac{12}{75} = 0.16$. The False Negative Rate measures the proportion of actual positive cases that are incorrectly classified as negative, and it is calculated as $\text{FNR} = \frac{\text{False Negative}}{\text{False Negative} + \text{True Positive}} = \frac{13}{13 + 39} = \frac{13}{52} = 0.25$. Comparing these values, the FNR is higher than the FPR. This means the model 4 is more likely to incorrectly classify a high-crime tract as low crime than to incorrectly classify a low-crime tract as high crime.

##### False Discovery Rate (FDR) vs False Omission Rate (FOR)

The False Discovery Rate measures the proportion of predicted positive cases that are actually negative and it is calculated as $\text{FDR} = \frac{\text{False Positive}}{\text{False Positive} + \text{True Positive}} = \frac{12}{12 + 39} = \frac{12}{51} \approx 0.235$. The False Omission Rate measures the proportion of predicted negative cases that are actually positive, and it is calculated as $\text{FOR} = \frac{\text{False Negative}}{\text{False Negative} + \text{True Negative}} = \frac{13}{13 + 63} = \frac{13}{76} \approx 0.171$. Comparing these values, the FDR is higher than the FOR, which means that the model 4 predicts that a census tract has high crime than median value, there is about 23.5% chance that this prediction is incorrect. When the model predicts that a tract has low crime rate than the median value, there is about 17.1% chance that the prediction is incorrect.

Overall, the model is somewhat more likely to miss high-crime areas, false negatives, than to incorrectly label low-crime areas as high crime, false positive. This suggests that the model tens to under-predict high crime rates, which could be an important consideration depending on how the model is used.
