---
title: "Outliers and Leverage"
author: "Dr. Josh Jacobson"
date: "18 February 2026"
output:
  pdf_document: default
---

```{r setup, include=FALSE,warning=FALSE,message=FALSE}
library(here)
library(knitr)
library(tinytex)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%", fig.align = "center")
```

*These lecture notes were originally developed by Dr. Kris Pruitt.*

The purpose of this lesson is to diagnose and resolve outliers within the training data for a regression model. Specifically, we distinguish between outliers with regard to the model's residuals and outliers among the predictor values.

# Types of Outliers

In the context of linear regression, outliers are observations that stray far from the rest of the training data. Visually, we could observe outliers in a scatter plot as observations that are distinct from the main "cloud" of points. If an observation is far from the rest of the data with regard to the fitted response variable value, then it will have a large residual (in absolute value). On the other hand, if an observation differs greatly from the rest of the data with regard to the explanatory variable (predictor) we say it has high **leverage**.

Outliers with large residuals and/or high leverage can sometimes affect the estimated slope of the regression line. We call such observations **influential** points. Influential points can be problematic if they mask the true association between the variables, so they warrant careful attention. We'll begin the lesson by exploring these concepts in scatter plots.

For this example, we will use data from a 1985 study regarding the body composition of males from ages 22-81. Import and review the file `bodyfat.csv` below.

```{r}
# import data
bodyfat <- read_csv("bodyfat.csv")

# review data structure
glimpse(bodyfat)
```

Imagine we are interested in the association between height (in inches) and weight (in pounds). As a result, we generate the following scatter plot and best-fit line.

```{r}
# display scatter plot
ggplot(data = bodyfat, aes(x = Height, y = Weight)) +
  geom_point(position = "jitter", alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE) +
  labs(
    x = "Height (in)",
    y = "Weight (lbs)"
  ) +
  theme_bw()
```

There are two clear outliers in this data that stray from the main "cloud" of points. With regard to the response variable (Weight), there is a very large value over 350 pounds. This observation has a very large residual, since 350+ pounds is far from what the fitted line would predict for a height of 72 inches. However, it is plausible for a six-foot tall male to have such a body weight so we have no reason to believe this observation is due to some kind of error. Consequently, we have no reason to remove it from the data set.

Regarding the explanatory variable (Height), there is a very small outlier below 30 inches. This observation has very high leverage, since its distance from the rest of the data is nearly 2.5 times the width of the main "cloud." This data point warrants some skepticism. A quick internet search reveals that an adult height below 30 inches (just over 2 feet) would be nearly a world record. Furthermore, for such an individual to weigh over 200 pounds seems almost physically impossible. This observation is likely an entry error. Rather than 29.5 inches, perhaps it should have been 69.5 inches? There is really no way for us to know without access to the authors of the study. However, this observation is so implausible that it could arguably be removed from the data set.

Despite it likely being an error, the outlier for height helps visually demonstrate why we use the term leverage for such points. In the context of physics, imagine the blue fitted line is a lever acting on the main "cloud" of points and that the outlier is the "person" using the lever. The outlier would have a lot of leverage in this context. This is a problem because the blue line does not appear to follow the trend displayed in the main "cloud" of points.

# Detection Methods

In the previous section, we investigated outliers and their influence from a purely visual standpoint. There are more objective methods for assessing the influence of an observation on the regression parameter estimates. **In general, an observation is considered influential if the fitted line when the point is excluded differs greatly from the fitted line when the point is included.** Let's fit models with and without the high leverage point and review the coefficient estimates.

```{r}
# fit model with outlier
model_with <- lm(Weight ~ Height, data = bodyfat)

coefficients(summary(model_with))
```

The estimated slope when we include the high leverage point is about 2.5. Now let's remove the high leverage point from the data and fit the new line.

```{r}
# remove outlier
bodyfat_filter <- bodyfat %>%
  filter(Height > 30)

# fit model without outlier
model_without <- lm(Weight ~ Height, data = bodyfat_filter)

coefficients(summary(model_without))
```

After removing the high leverage point, the estimated slope increases dramatically to about 5.5. To put this difference in context, let's think about the proper interpretation of the estimated slope. We expect a one inch increase in height to be associated with a 5.5 pound increase in weight, on average. That's more than double the average weight increase compared to the estimate with the high leverage point! In general, we don't want a single observation to have such an influence on our estimates and insights. In this case, because we already think the observation is an error, the easy answer is to remove it. However, the answer isn't always so easy in real-world applications.

Now let's conduct the same analysis for the large residual point. What impact does its removal have on the estimated slope? For clarity, we leave the high leverage point out as well.

```{r}
# remove outlier
bodyfat_filter2 <- bodyfat %>%
  filter(Height > 30, Weight < 350)

# fit model without outlier
model_without2 <- lm(Weight ~ Height, data = bodyfat_filter2)

coefficients(summary(model_without2))
```

By comparison, the estimated slope changed very little. After removing the large residual point, the estimate decreased to 5.3 pounds per inch. Consequently, we would *not* label this observation as influential. Whether or not it is in the model, the estimated slope is roughly the same.

# Diagnostic Plots

Rather than manually fitting models with and without suspected influential outliers, we can rely on built-in diagnostic plots. Specifically, the Residuals versus Leverage plot provides exactly what we require.

```{r}
# display diagnostic plot
plot(model_with, which = 5)
```

The first thing that stands out in this plot is the red line pointing to a point labelled 42. This is the 42nd observation in the data frame. What is that observation?

```{r}
# display 42nd observation
bodyfat[42, c(4, 5)]
```

Observation 42 is the height outlier. This point has a high leverage value on the $x$-axis and a high residual value on the $y$-axis. As a result, it appears in the upper-right corner of the plot. In general, when points appear in the upper-right or lower-right corners of the plot, they are candidates for high influence. In the upper-left corner of the diagnostic plot, we see observation 39. Which observation is this?

```{r}
# display 39th observation
bodyfat[39, c(4, 5)]
```

That is our weight outlier. By contrast, this observation has a high residual value on the $y$-axis, but a low leverage value on the $x$-axis. Observations in the upper-left or lower-left of the plot have very low leverage and are seldom labelled as influential. Rather than subjectively assessing the location of points on the plot, we employ a more objective metric called the **Cook's Distance**.

The Cook's distance is a measure of how influential a data point is based on the impact of removing it from the model. In a sense, the Cook's distance mimics what we did in the previous section and assigns a value to the differences between the two models (with and without). Larger residuals and higher leverage will both increase the Cook's distance. Points beyond a Cook's distance of 1 (second dashed line) are generally considered highly influential. Thus, we have further confirmation that the height outlier is highly influential and the weight outlier is not.

In terms of the mathematics supporting this diagnostic plot, the leverage ($h_i$) for observation $i$ is the $i$-th diagonal element of the $H$ matrix below. For a model with $n$ observations, the leverage will always be between $1/n$ and 1.

$$
H=X(X^TX)^{-1}X^T
$$

This matrix should look very familiar as part of the least-squares solution to find the predicted values for the response $\hat{y}$. We then use the leverage of observation $i$, along with its squared residual $\epsilon_i^2$, to determine the Cook's Distance for a model with $p$ predictors.

$$
D_i=\frac{\epsilon_i^2}{p \cdot \text{MSE}} \cdot \frac{h_i}{(1-h_i)^2}
$$

Notice that a larger residual or larger leverage result in a greater Cook's Distance.
