---
title: "Multicollinearity"
author: "Dr. Josh Jacobson"
date: "20 February 2026"
output:
  pdf_document: default
---

```{r setup, include=FALSE,warning=FALSE,message=FALSE}
library(here)
library(knitr)
library(tinytex)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%", fig.align = "center")
```

*These lecture notes were originally developed by Dr. Kris Pruitt.*

The purpose of this lesson is to assess the impact of multicollinearity in multiple linear regression models. In particular, we will identify multicollinearity (visually and analytically) and measure its effect via the variance inflation factor.

# Identifying Multicollinearity

When two predictor variables are correlated with one another, we say they are **collinear**. A model with multiple instances of correlated predictor variables is said to suffer from **multicollinearity**. Often we see these two terms used interchangeably, but the statistical issues are the same. When two predictor variables are correlated with one another, it is difficult to isolate the true effect that a single predictor has on the response. In some sense, including both collinear variables is like "double-counting" the impact on the response. We'll begin the lesson by identifying collinearity and then move on to evaluating its effect.

For this example, we will return to the body composition data found in the file `bodyfat.csv` on the course site. Import and wrangle the data using the code below.

```{r}
# import and wrangle data
bodycomp <- read_csv(here("data/bodyfat.csv")) %>%
  filter(Height > 30, Weight < 350) %>%
  transmute(Weight, Height,
    Chest = Chest / 2.54,
    Abdomen = Abdomen / 2.54
  ) %>%
  na.omit()
```

Imagine we are interested in explaining a person's weight (lbs) based on their height (in), as well as their chest and abdomen circumferences (in). We can visually check for collinearity using a matrix of scatter plots.

```{r}
# create scatter plot matrix
library(GGally)
ggpairs(
  bodycomp,
  upper = list(continuous = wrap("points")),
  lower = list(continuous = wrap("cor"))
)
```

The scatter plots reveal good news and bad news! The good news is that all three predictors appear to have a positive, linear association with the response. This makes intuitive sense because we would expect taller, larger people to weigh more. The bad news is that chest and abdominal circumference appear to be strongly correlated with one another. This also makes intuitive sense, but it complicates any inferences we might want to make regarding the influence on weight. We can similarly check for collinearity using a correlation matrix.

```{r}
# create correlation matrix
cor(bodycomp)
```

The numerical results align with our visual results. The positive linear association with the response is moderate for height and strong for chest and abdomen circumferences. Height has no clear association with the circumferences, so we have no concerns for collinearity with that predictor. However, there is a very strong association between chest and abdomen circumferences. Consequently, we would not want to use both of those explanatory variables in an inferential model of weight.

# Evaluating the Impact

Once we've identified collinear pairs of variables, we can evaluate the impact of including both in an inferential model. From a mathematical standpoint, the issue regards the standard error for the slope coefficients. Recall, we use the following formula to construct a confidence interval for each slope parameter in a multiple linear regression model.

$$
\beta_i \in \hat{\beta}_i \pm t_{\alpha/2} \cdot SE_{\hat{\beta}_i}
$$

For a given parameter $\beta_i$, the standard error $SE_{\hat{\beta}_i}$ is greater when a collinear variable exists in the model. As a result, the margin of error is greater than it should be, even at the same level of confidence. We'll continue the lesson by demonstrating this phenomenon in an example.

In the previous section, we saw that chest and abdomen circumferences are strongly correlated. So, we do not want to keep both in the regression model. But, which one should we keep? One simple rule is to keep the variable that is most strongly correlated with the response. In our case, that is the chest circumference variable. There are more sophisticated variable selection methods than this, but the correlation-based method is sufficient for our demonstration here. Let's fit the model of weight with all three explanatory variables and determine the standard error for the chest circumference estimate.

```{r}
# fit model with abdomen
model1 <- lm(Weight ~ Height + Chest + Abdomen, data = bodycomp)

# extract coefficient statistics
coefficients(summary(model1))
```

The standard error for the variable we want to keep in the model (i.e., chest circumference) is 0.358. Now, let's compute the standard error when the collinear variable (i.e., abdomen) is removed.

```{r}
# fit model without abdomen
model2 <- lm(Weight ~ Height + Chest, data = bodycomp)

# extract coefficient statistics
coefficients(summary(model2))
```

The standard error for chest circumference decreased to 0.176. This is half of the standard error from before! Let's compare the margin of error ($\text{MOE} = t_{\alpha/2} \cdot SE_{\hat{\beta}_i}$) for 95% confidence intervals in both cases.

```{r}
# compute margin of error with abdomen
qt(p = 0.025, df = nrow(bodycomp) - 4, lower.tail = FALSE) * 0.358

# compute margin of error without abdomen
qt(p = 0.025, df = nrow(bodycomp) - 3, lower.tail = FALSE) * 0.176
```

The margin of error for a confidence interval on the chest circumference coefficient is 0.705 pounds per inch when abdomen is included in the model. Without abdomen in the model, the margin of error is reduced to 0.347. Thus, removing the collinear variable improves the precision of our estimate for the effect of chest circumference on body weight.

For purposes of comparison, observe the difference in standard error for height between the two models. There is much less of a difference (0.185 versus 0.215) with or without abdomen in the model. This is a strong indicator that height is *not* collinear with abdomen, because the existence of abdomen in the model has little impact on the estimate of standard error for height.

# Variance Inflation Factor

A more comprehensive method to check for multicollinearity among *all* the variables in the model is to compute a **variance inflation factor (VIF)**. The VIF measures the total increase in standard error across all variables when another potentially-collinear variable is included in the model. For the $i$th explanatory variable, the VIF is calculated as:

$$
VIF=\frac{1}{1-R_i^2}
$$

where $R_i^2$ is the R-squared value obtained by fitting a regression model with the $i$th variable as the response and the remaining ones as explanatory variables. If $R_i^2$ is large, then the $i$th explanatory variable is highly correlated with some or all of the others. A large $R_i^2$ results in a relatively large $VIF$ and an indication of multicollinearity. There is no agreed-upon rule for how large of a $VIF$ is *too* large. However, a common threshold is a value of 5.

Let's demonstrate VIF by fitting a model with abdomen as the response and with height and chest circumference as explanatory variables.

```{r}
# fit model
model3 <- lm(Abdomen ~ Height + Chest, data = bodycomp)

# extract R-squared
summary(model3)$r.squared
```

The model indicates $R_i^2=0.829$ when abdomen is treated as the $i$th explanatory variable. Using the formula above, this is equivalent to $VIF=5.85$. Because this $VIF$ value exceeds 5, we have concern for multicollinearity in the model. Since $VIF$ is connected to one explanatory variable at a time, we would need to repeat this process for every variable. Luckily, there are packages for `R` that complete this repetitive task for us.

```{r}
# load car package
library(car)

# compute VIF for all predictors
vif(model1)
```

The output of the `vif()` function from the `car` package is the same $R_i^2$ calculation we completed manually. As we've suspected all along, there is no issue with multicollinearity regarding the height predictor. But there *are* issues for both chest and abdomen circumferences because they are collinear with one another. So, one of them should be removed from the model if we hope to obtain accurate confidence intervals on the slope parameters.

Because the primary impact of multicollinearity is the inflation of variability surrounding the slope parameter estimates, it is most worrisome when conducting inference. If our primary focus is on *prediction*, then there is less of a concern for multicollinearity. This is due to differences in performance metrics. When conducting inference, we want precise estimates of the slope parameters (i.e., small margin of error). When pursuing prediction, we only care about the accuracy of the estimates for the response variable. So, if a collinear model is more accurate, in terms of prediction, we may choose to ignore the variance inflation.
