---
title: "Multiple Logistic Regression"
author: "Dr. Josh Jacobson"
date: "4 March 2026"
output:
  pdf_document: default
---

```{r setup, include=FALSE,warning=FALSE,message=FALSE}
library(here)
library(knitr)
library(tinytex)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%", fig.align = "center")
```

*These lecture notes were originally developed by Dr. Kris Pruitt.*

The purpose of this lesson is to fit a multiple logistic regression model using the maximum likelihood method. We will leverage built-in functions in `R` to execute the algorithm and to measure the fit of the resulting models.

# Maximum Likelihood Method

The methodology for estimating a multiple logistic regression model is very similar to that of the single-predictor (simple) version. The difference is that we now wish to predict a binary response $y$ based on $k>1$ predictor variables $x_1,x_2,...,x_k$. The population model is described by the following function.

$$
p(x)=P(y=1|x)=\frac{1}{1+e^{-(\beta_0+\beta_1x_1+\beta_2x_2+\cdots+\beta_kx_k)}}
$$

This function suggests an underlying logistic relationship between the vector of explanatory variables ($x$) and the binary response. Just as with simple logistic regression, we seek estimates $\hat{\beta}_0,\hat{\beta}_1,...,\hat{\beta_k}$ that provide the maximum log-likelihood value (shown below). In this formula, $\hat{p}_i$ represents the estimated probability $p(x)$ for each observation $i$.

$$
\text{log(L)}=\sum_{i=1}^n [y_i \log(\hat{p}_i)+(1-y_i)\log(1-\hat{p}_i)]
$$

Because the maximum likelihood is determined via numerical methods, we leverage built-in functions in `R` for the computation. Let's demonstrate a multiple logistic regression model in an example.

# Multiple Logistic Regression

We return to the cancer cell imagery example where our goal is to determine the probability that a breast cancer cell is malignant. Import, wrangle, and review the data below.

```{r}
# import and wrangle data
cancer <- read_csv("cancer.csv") %>%
  transmute(
    Diagnosis = ifelse(diagnosis == "M", 1, 0),
    Radius = radius_mean,
    Texture = texture_mean,
    Smoothness = smoothness_mean,
    Symmetry = symmetry_mean
  ) %>%
  na.omit()

# review data
glimpse(cancer)
```

We have four potential explanatory variables in the data. As with linear regression, a good first step is to construct a matrix of scatter plots and look for evidence of association.

```{r}
# display scatter plot matrix
pairs(cancer)
```

Each explanatory variable appears to be positively associated with the response and diagnosis seems to be the strongest. Additionally, smoothness and symmetry appear to be collinear. We can check all of the pair-wise correlation coefficients in the matrix below.

```{r}
# compute correlation coefficients
cor(cancer,
  method = "pearson"
)
```

As expected, radius has the strongest correlation with diagnosis and smoothness has a moderate correlation with symmetry. The collinearity between these two explanatory variables may not be strong enough to significantly impact the model, but we should be aware of its potential impact on our standard error estimates. Let's fit a model for diagnosis with cell radius and texture.

```{r}
# fit logistic regression model
model1 <- glm(Diagnosis ~ Radius + Texture, data = cancer, family = "binomial")

# display summary output
summary(model1)
```

Based on the null and residual deviance, we have a pseudo-R-squared value of 0.61. Thus, we have eliminated 61% of the deviance from the perfect (saturated) model. In log-odds form, the estimated model is:

$$
\log\left(\frac{P(\text{Malignant})}{P(\text{Benign})}\right)=-19.849+1.057\cdot\text{Radius}+0.218\cdot\text{Texture}
$$

We know from the previous lesson's likelihood ratio test that at least one of these predictors is significant. Based on their individual p-values, it appears *both* slope coefficients significantly differ from zero. All told, we seem to have a model that fits the data well. However, we may be able to improve the fit by adding more explanatory variables. Let's try adding smoothness.

```{r}
# fit logistic regression model
model2 <- glm(Diagnosis ~ Radius + Texture + Smoothness, data = cancer, family = "binomial")

# display summary output
summary(model2)
```

Our residual deviance for the new model is much smaller. But if we want to compare two models of different sizes, then we need to use AIC. Recall that AIC penalizes the deviance for adding more explanatory variables. Even with the penalty, the AIC for the new model (195.29) is smaller than that of the old model (297.12). Might we improve the model even more by adding symmetry?

```{r}
# fit logistic regression model
model3 <- glm(Diagnosis ~ ., data = cancer, family = "binomial")

# display summary output
summary(model3)
```

The AIC is a bit smaller (189.65 vs. 195.29), but the decrease is not nearly as large as what we achieved by adding smoothness. The p-value on symmetry is significant, but not nearly as small as that of the other variables. All of this is likely due to symmetry and smoothness being collinear. When we already have smoothness in the model, adding symmetry offers relatively little value. That said, the small increase in the standard error on smoothness indicates that the collinearity is not incredibly strong. So, it is probably safe to keep symmetry in the model and accept the minor improvement in fit. Ultimately, we have eliminated 76% of the deviance from the perfect model.

# Interactions Between Predictors

Just as with multiple linear regression, we can include categorical predictors or interaction terms in our model, when appropriate. For example, a model with two predictor variables $x_1$ and $x_2$ could include the multiplicative term $x_1x_2$ in the following logistic function.

$$
p(x)=P(y=1|x)=\frac{1}{1+e^{-(\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2)}}
$$

In log-odds form, the equivalent model can be written as follows:

$$
\log\left(\frac{P(y=1|x)}{P(y=0|x)}\right)=\beta_0+\beta_1x_1+(\beta_2+\beta_3x_1)x_2
$$

After the linear transformation, and some algebra, it is clear that the influence of $x_2$ on the log-odds is affected by the value of $x_1$. Let's investigate this concept using the example data.

```{r}
# fit logistic regression model
model4 <- glm(Diagnosis ~ Radius * Texture, data = cancer, family = "binomial")

# display summary output
summary(model4)
```

In this example, $x_1$ is the cell radius and $x_2$ is the cell texture. We've added an interaction term to see if the influence of texture on diagnosis is affected by radius. Although the new interaction term is significant at the $\alpha=0.05$ level of significance, there is very little improvement in the AIC (294.06 versus 297.12) compared to the model with only the main effects. So, there is little evidence of a strong interaction. Let's try including two-way interactions between all of the predictors.

```{r}
# fit logistic regression model
model5 <- glm(Diagnosis ~ .^2, data = cancer, family = "binomial")

# display summary output
summary(model5)
```

Notice we used the shortcut `.^2` to add all of the individual predictors *and* all of the pairwise interactions between them. Here, we obtain an interesting result. The AIC actually got *worse* (190.10 versus 189.65) compared to the model with only the individual predictors. So, the reduction in deviance was not enough to overcome the penalty for additional predictors. Perhaps we should only include some of the interactions? We'll learn techniques for selecting the right combination of predictors later in the course.
